\mytitle

### 2021-12-27 paper.tex
fix typos

As is well known, the performance of this naive implementation can be
improved by applying Horner's scheme \cite{horner}, which reduces the
number of expensive floating\-/point multiplications to 7 and requires
6 additions:
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \bigg\{1 + x \bigg[-\frac{1}{2} + x \bigg(\frac{1}{3} + x \bigg\{-\frac{1}{4} \\
  &+ x \bigg[\frac{1}{5} + x\bigg(-\frac{1}{6} + \frac{x}{7}\bigg)\bigg]\bigg\}\bigg)\bigg]\bigg\}.
\end{split}\label{eq:horner}%
\end{align}
%
A C implementation using Horner's scheme could read
%
\lstinputlisting[language=C]{src/log1p_horner.c}
%
Note, that the formulation using Horner's scheme requires the
evaluation of the floating\-/point operations in strictly sequential
order.  However, the significant reduction of the number of
floating\-/point multiplications, compared to the naive
implementation, usually results in a shorter run\-/time.

There are several different possibilities to reduce the run\-/time of
Horner's method by rewriting the series expansion of $\ln(1+x)$ such
that more terms can be calculated in parallel, while the number
floating\-/point operations is still kept small.  One possibility is
to split the r.h.s.\ of Eq.~\eqref{eq:horner} into two distinct sums,
one containing the even powers and the other containing the odd powers
of $x$,
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \left\{1 + y \left[\frac{1}{3} + y
  \left(\frac{1}{5} + \frac{y}{7}\right)\right]\right\} \\
  &- y\left[\frac{1}{2} + y\left(\frac{1}{4} + \frac{y}{6}\right)\right],
\end{split}\label{eq:split}%
\end{align}
%
where $y=x^2$.  This formulation requires 8 floating\-/point
multiplications and 6 additions, i.e.\ one more multiplication than
Horner's scheme.  However, in Eq.~\eqref{eq:split} both sums can then
be calculated in parallel, leading to a potential speed-up if the cost
of the additional multiplication is smaller than the gain by the
parallel execution.  A C implementation could read:
%
\lstinputlisting[language=C]{src/log1p_split.c}
%
Another possibility is to rewrite the r.h.s.\ of Eq.~\eqref{eq:taylor}
using Estrin's scheme \cite{estrin},
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x + \left(\frac{x}{3}-\frac{1}{2}\right)
    y \\ &+\left[\left(\frac{x}{7}-\frac{1}{6}\right)
    y+\frac{x}{5}-\frac{1}{4}\right] z,
\end{split}\label{eq:estrin}%
\end{align}
%
where $y=x^2$ and $z=y^2$.  This form also requires 8 floating\-/point
multiplications and 6 additions, as in Eq.~\eqref{eq:split}.  However,
in Estrin's form more terms can be executed in parallel, leading to a
potential further speed-up.  A C implementation using Estrin's scheme
could read:
%
\lstinputlisting[language=C]{src/log1p_estrin.c}

To exploit ILP further, one could use a rational function
approximation, such as a Padé approximant \cite{pade}, where the
numerator and the denominator can be calculated in parallel.  A Padé
approximant of $\ln(1+x)$ at the same order as Eq.~\eqref{eq:taylor}
reads
%
\begin{align}
  \ln(1+x) &\approx x
  \frac{\frac{320 x^3}{40143}+\frac{44320 x^2}{120429}+\frac{51064
    x}{40143}+1}{\frac{5120 x^3}{40143}+\frac{12320 x^2}{13381}+\frac{23712
    x}{13381}+1}
  \label{eq:pade}
  \\
  &= x
  \frac{\left(\frac{320 x}{40143}+\frac{44320}{120429}\right)y+\frac{51064
    x}{40143}+1}{\left(\frac{5120 x}{40143}+\frac{12320}{13381}\right)y+\frac{23712
    x}{13381}+1},
  \label{eq:mixed}
\end{align}
%
where $y=x^2$.  In Eq.~\eqref{eq:mixed} the numerator and the
denominator have been re-written using Estrin's scheme, allowing for
further ILP.  A C implementation of Eq.~\eqref{eq:mixed} could read:
%
\lstinputlisting[language=C]{src/log1p_mixed.c}
%
Note, that the value $y=x^2$ can be re-used in the calculation of both
the numerator and the denominator.  A computation of $y^2$ is not
necessary.  This ``mixed'' implementation requires 8 floating\-/point
multiplications, 1 floating\-/point division and 6 additions.  The
rational function approximation in Eq.~\eqref{eq:mixed} may be
beneficial for long numerator/denominator polynomials, where the cost
of the floating\-/point division is smaller than the gain by the
parallel execution of the terms in the numerator and denominator.

>>>>

As is well known, the performance of this naive implementation can be
improved by applying Horner's scheme \cite{horner}, which reduces the
number of expensive floating\-/point multiplications to 7 and requires
6 additions:
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \bigg\{1 + x \bigg[-\frac{1}{2} + x \bigg(\frac{1}{3} + x \bigg\{-\frac{1}{4} \\
  &+ x \bigg[\frac{1}{5} + x\bigg(-\frac{1}{6} + \frac{x}{7}\bigg)\bigg]\bigg\}\bigg)\bigg]\bigg\}.
\end{split}\label{eq:horner}%
\end{align}
%
A C implementation using Horner's scheme could read
%
\lstinputlisting[language=C]{src/log1p_horner.c}
%
Note, that the formulation using Horner's scheme requires the
evaluation of the floating\-/point operations in strictly sequential
order.  However, the significant reduction of the number of
floating\-/point multiplications, compared to the naive implementation
from above, usually results in a shorter run\-/time.

There are several different possibilities to reduce the run\-/time of
Horner's method by rewriting the series expansion of $\ln(1+x)$ such
that more terms can be calculated in parallel, while the number
floating\-/point operations is still kept small.  One possibility is
to split the r.h.s.\ of Eq.~\eqref{eq:horner} into two distinct sums,
one containing the even powers and the other containing the odd powers
of $x$,
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \left\{1 + y \left[\frac{1}{3} + y
  \left(\frac{1}{5} + \frac{y}{7}\right)\right]\right\} \\
  &- y\left[\frac{1}{2} + y\left(\frac{1}{4} + \frac{y}{6}\right)\right],
\end{split}\label{eq:split}%
\end{align}
%
where $y=x^2$.  This formulation requires 8 floating\-/point
multiplications and 6 additions, i.e.\ one more multiplication than
Horner's scheme.  However, in Eq.~\eqref{eq:split} both sums can then
be calculated in parallel, leading to a potential speed-up if the cost
of the additional multiplication is smaller than the gain by the
parallel execution.  A C implementation could read:
%
\lstinputlisting[language=C]{src/log1p_split.c}
%
Another possibility is to rewrite the r.h.s.\ of Eq.~\eqref{eq:taylor}
using Estrin's scheme \cite{estrin},
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x + y \left(\frac{x}{3}-\frac{1}{2}\right)
    \\ &+ z \left[\frac{x}{5} - \frac{1}{4}
      + y \left(\frac{x}{7}-\frac{1}{6}\right)
    \right],
\end{split}\label{eq:estrin}%
\end{align}
%
where $y=x^2$ and $z=y^2$.  This form also requires 8 floating\-/point
multiplications and 6 additions, as in Eq.~\eqref{eq:split}.  However,
in Estrin's form more terms can be executed in parallel, leading to a
potential further speed-up.  A C implementation using Estrin's scheme
could read:
%
\lstinputlisting[language=C]{src/log1p_estrin.c}

Another option to exploit ILP is to use a rational function
approximation, such as a Padé approximant \cite{pade}, where the
numerator and the denominator can be calculated in parallel.  A Padé
approximant of $\ln(1+x)$ at the same order as Eq.~\eqref{eq:taylor}
can be written as
%
\begin{align}
  \ln(1+x) &\approx x
  \frac{
    1 + \frac{51064}{40143} x + \frac{44320}{120429} x^2 + \frac{320}{40143} x^3
  }{
    1 + \frac{23712}{13381} x + \frac{12320}{13381} x^2 + \frac{5120}{40143} x^3
  }
  \label{eq:pade}
  \\
  &= x
  \frac{
    1 + \frac{51064}{40143} x + y \left(\frac{44320}{120429} + \frac{320}{40143} x\right)
  }{
    1 + \frac{23712}{13381} x + y \left(\frac{12320}{13381} + \frac{5120}{40143} x\right)
  },
  \label{eq:mixed}
\end{align}
%
where $y=x^2$.  In Eq.~\eqref{eq:mixed} the numerator and the
denominator have been re-written using Estrin's scheme, allowing for
further ILP, compared to Eq.~\eqref{eq:pade}.  A C implementation of
Eq.~\eqref{eq:mixed} could read:
%
\lstinputlisting[language=C]{src/log1p_mixed.c}
%
Note, that the value $y=x^2$ can be re-used in the calculation of both
the numerator and the denominator.  A computation of $y^2$ is not
necessary.  This ``mixed'' implementation requires 8 floating\-/point
multiplications, 1 floating\-/point division and 6 additions.  The
rational function approximation in Eq.~\eqref{eq:mixed} may be
beneficial for long numerator/denominator polynomials, where the cost
of the floating\-/point division is smaller than the gain by the
parallel execution of the terms in the numerator and denominator.
